---
title: "ggplot continued, loops"
author: "Sean Harrington"
date: "2023-03-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/r4grads/Fish_data/Modified/")
```

<br>

[Home](https://wyoibc.github.io/r4grads/)

<br>
<br>

## Set up:

If you don't have the data from last time handy, or have forgotten where you put it, use the instructions from the the [week 2 session](https://wyoibc.github.io/r4grads/Module_2/R_Data_manip.html) to re-download it. Then set your working directory to where the data are:

* **Remember that this file path will almost definitely be different for you!**

```{r, eval = FALSE}
setwd("~/r4grads/Fish_data/Modified/")
```

We'll need our `all_data` object again. If you have your `"fish_data_merged.csv"` file that we wrote out last time, you can read in:

```{r}
all_data <- read.csv("fish_data_merged.csv")
```

If you've lost track of this file, we can create it again:

* **If you have the "fish_data_merged.csv" file and were able to run the line above, you do not need to run these lines!**

```{r}
# read in each of the data files
body <- read.csv("Fish_body_size.csv")
iso <- read.csv("Fish_isotopes.csv")
# Make all of the subsitutions that we made last time
body$Species <- gsub("Steelhesd", "Steelhead", body$Species)
body$Species <- gsub("COHO", "Coho", body$Species)
body$Species <- gsub("^Dolly$", "Dolly varden", body$Species)
body$Site <- gsub("RT02-R", "RT02R", body$Site)
body$Site <- gsub("RT02-BP", "RT02BP", body$Site)
# Merge the data
all_data <- merge(body, iso)
```

<br>
<br>

# ggplot continued

<br>
<br>

Load up tidyverse:

```{r}
library(tidyverse)
```


<br>


### Geometric objects and facets

There are many other types of geoms, or geometric objects, that we can plot. We won't come anywhere close to exploring the full set that are available, but we'll demonstrate a few others.


Here is a smoothed line plot of length vs. del15N:

```{r}
ggplot(data = all_data) + 
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N))
```

<br>

We can show the different species by setting `linetype`:

```{r}
ggplot(data = all_data) + 
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N, linetype = Species))
```

This does not look good at all and throws some warnings These are caused because at least one of our species has very few observations, and this causes issues with the method used to create the smoothed line.

* **Warnings vs. errors**: An error in R means that something has gone wrong such that the function you tried to run could not be successfully executed. Warnings are *potentially* less severe. It means that some potetentially undesirable behavior has occurred, but that the function is still able to run and return output. However, a warning could very well mean that your function has returned bad results, and you should always investigate what caused a warning and if the output is acceptable or not.

<br>

In this case, it might be useful to look at each species plotted individually. However, I don't want to go through and individually code each plot. Instead, let's use the facet functionality of ggplot:


```{r}
ggplot(data = all_data) + 
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N)) +
  facet_wrap(~ Species)
```

We can start to see where the problem might be, but the x and y scales for these different species really aren't comparable, so let's allow those to change freely:


```{r, warning = FALSE}
ggplot(data = all_data) + 
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N)) +
  facet_wrap(~ Species, scales = "free")
```

As is the case with most things in `ggplot` and R more generally, there are many ways to customize facets, and we won't explore those here.

* Which of these species looks problematic? Try making a single plot using `geom_smooth()` including `linetype = Species` as above, but remove the offending species.


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

I removed only "Pink" - you might have decided to also remove "Chum" - either is fine.


Using base R indexing:

```{r, fig.show = 'hide', message = FALSE}
ggplot(data = all_data[!all_data$Species == "Pink", ]) + 
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N,  linetype = Species))
```


Using filter from `dplyr`:

```{r}
ggplot(data = filter(all_data, Species != "Pink")) + # Note the != for "not equal to"
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N,  linetype = Species))
```

This works, but, it's still rather hard to tell what's going on with each species. 

* Try adding an argument to make the lines different colors.


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
ggplot(data = filter(all_data, Species != "Pink")) +
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N,  linetype = Species, color = Species))
```
<br>
<br>


This looks reasonably good, but still a little busy. This may or may not be the best way to convey these data. For the next few things we try, let's reduce out dataset down to just Coho, coastrange sculpin, and Steelhead.

```{r}
data_red <- filter(all_data, Species %in% c("Coho", "coastrange sculpin", "Steelhead"))
```

<br>
<br>

We will often want to stack multiple geoms on top of each other. E.g., to fully visualize the trends and spread in the data, we might want to plot both the points and the line:


```{r}
ggplot(data = data_red) + 
  geom_point(mapping = aes(x = Fork.length..cm., y = del15N)) +
  geom_smooth(mapping = aes(x = Fork.length..cm., y = del15N))
```


<br>

If we have multiple layers in our plot that all share the same aesthetic, it can be easier to specify them once in the `ggplot()` call rather than idividually each time we call a new geom:

```{r}
ggplot(data = data_red, mapping = aes(x = Fork.length..cm., y = del15N)) + 
  geom_point() +
  geom_smooth()
```

<br>

We get exactly the same plot. The global mappings are applied to each geom when it is called. We can also extend or override the global mappings within each geom:

```{r}
ggplot(data = data_red, mapping = aes(x = Fork.length..cm., y = del15N)) + 
  geom_point(mapping = aes(color = Species)) +
  geom_smooth()
```

<br>
<br>

* Make a plot that has both points colored by species, including separate colored lines for each species as well.


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
ggplot(data = data_red, mapping = aes(x = Fork.length..cm., y = del15N, color = Species)) + 
  geom_point() +
  geom_smooth(mapping = aes(linetype = Species))
```


<br>

Note that we can also add the data argument into a geom. Here we'll plot the points for all species, but the line for only Coho:

```{r}
ggplot(data = data_red, mapping = aes(x = Fork.length..cm., y = del15N, color = Species)) + 
  geom_point() +
  geom_smooth(data = filter(data_red, Species == "Coho"))
```


There are many, many ways to futher customize graphics in ggplot. If there is a type of plot you want to make, there is a good chance that ggplot will make it and you can usually find good examples with some internet searching. I turn to Google all the time to figure out how to make specific plots in ggplot.

<br>
<br>

### Plots by discrete variables

We've made a bunch of scatter and line plots, let's explore a few ways to plot out discrete variables.

<br>

We can make just a scatterplot, but it's really not very informative.

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_point()
```


<br>
<br>

We can make a boxplot instead:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_boxplot()
```

This tells us a lot more information. We can also do things like swap the x and y coordinates:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_boxplot() +
  coord_flip()
```

Or fill the boxes with colors by species:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N, fill = Species)) + 
  geom_boxplot() +
  theme(legend.position="none")
```

* What does `theme(legend.position="none")` in the above code do? What happens if we remove it?

<br>
<br>
<br>
<br>

There are lots of other things we could do with boxplots, but we'll move on. My favorite way of visualizing distributions for discrete variables is using violin plots:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_violin(trim=FALSE)
```


This is fine, but not great. Change this plot so that the violins are filled with different colors for each species. Also get rid of any legend that shows up. 



<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N, fill = Species)) + 
  geom_violin(trim=FALSE) +
  theme(legend.position="none")
```

<br>

If we want to, we can add boxplots within the violins. This is similar to adding the `geom_smooth` to the scatterplots above. Try making a violin plot with boxplots inside. 

Make the violins filled by different colors for each species, but keep the boxplots unfilled. As a hint, the argument `width` controls the width of boxplots.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_violin(trim=FALSE, mapping = aes(fill = Species)) +
  geom_boxplot(width=0.1)+
  theme(legend.position="none")
```


This looks pretty fancy to me.

<br>
<br>

A couple final things:

We can change the axis labels and add a plot title if we want:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_violin(trim=FALSE, mapping = aes(fill = Species)) +
  geom_boxplot(width=0.1)+
  theme(legend.position="none") +
  labs(title="Fish nitrogen isotopes", x = "Species", y = "delta 15 N")
```

And/or we can change the theme of the background:

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_violin(trim=FALSE, mapping = aes(fill = Species)) +
  geom_boxplot(width=0.1)+
  labs(x = "Species", y = "delta 15 N") +
  theme_minimal() +
  theme(legend.position="none")
```

* Note that we had to move `theme(legend.position="none")` to after `theme_minimal()` because whichever comes last will override some of the parameters set by the one that is run first.

or 

```{r}
ggplot(data = all_data, mapping = aes(x = Species, y = del15N)) + 
  geom_violin(trim=FALSE, mapping = aes(fill = Species)) +
  geom_boxplot(width=0.1)+
  labs(x = "Species", y = "delta 15 N") +
  theme_classic() +
  theme(legend.position="none")
```


<br>
<br>
<br>
<br>

You customize each of these plots almost endlesslessly, and there are many other types of plots that you can create, as well. If you want to get an idea for some of the plotting diversity available, check out the the [R graph gallery](https://r-graph-gallery.com/).




<br>
<br>

[Home](https://wyoibc.github.io/r4grads/)

<br>
<br>



